The state of knowledge for marine invertebrate biodiversity in the continental US

### 2017-07-06 manuscript.Rmd
more text edits

Untangling the relative contribution of these factors to explain the discrepencies is difficult as it would require to know the criteria used by the experts who have assembled the lists. However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepency, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico.

@@

We investigated the proportion of species discovered since the publication of the faunistic lists using iDigBio records. They represent `r sum(recently_collected$gom_in_idig$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_idig$min_year > 1987)` species for the Pacific Northwest. While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the presence of these species based only on museum records, as these lists were mostly assembled from publications, and only a small fraction of museum records are used in publications.

Museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000]. Some of the discrepancies observed between the number of species for which occurrence data and museum specimens are available suggest that additional digitization of collections (including capturing their geographic information) are needed. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase when records from the Florida Wildlife Research Institute collection (St Petersburg, FL) become integrated into the biodiversity information systems.

@@

We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected in the EEZ waters of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species. Most samples were collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A). For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline. For the West coast, the Pacific Northwest, the San Francisco Bay Area, and Southern California have the highest richness (Fig. \@ref(fig:map-diversity-sampling)B).

@@

These patterns of spatial distribution of the species richness are strongly confounded by the sampling effort. When investigating the relation between the number of samples and the number of species for each cell of the map (Fig. \@ref(fig:correlation-sampling)), we can see that until reaching about 100 records per cell there is little redundancy in the sampled species (each record represent a different species).

@@

The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that only `r prop_singleton_species`% of the species are known from a single record. The strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new observations (Fig. \@ref(fig:map-diversity-sampling)) and new species. Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century lead to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970's when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections after the 1970s has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). Therefore, coordinated large-scale biodiversity expeditions will continue to generate a wealth of data and new records. For instance, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014].

```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records. (\\#fig:idigbio-through-time)'}
knitr::include_graphics("figures/idigbio_through_time.pdf")
```

@@

To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time. For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens remain for the most part unidentified in collections \@ref{fig:identification-through-time}. For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%. The number of scientists describing new species might be increasing globally [@Costello2013], but local experts are needed to keep identifying the species that get collected.

>>>>

Untangling the relative contribution of these factors to explain the discrepencies is difficult as it would require to know the criteria used by the experts who have assembled the lists. However, by querying iDigBio we found that for the `r format(not_really_in_database$idig_gom_n_spp_not_in_db, big.mark=",")` species that do not appear within the geographic range of the Gulf of Mexico but are on the list, `r format_output(not_really_in_database$idig_gom_p_spp_not_in_db*100)`% are truly absent from iDigBio, while for the `r format(not_really_in_database$idig_gom_n_spp_no_coords, big.mark=",")` species that appear in iDigBio, `r format_output(not_really_in_database$idig_gom_p_spp_no_coords*100)`%  do not have geographic coordinates associated with them, and for the ones that do have geographic coordinates, `r format_output(not_really_in_database$idig_gom_p_spp_within_usa*100)`% ($n=$  `r format(not_really_in_database$idig_gom_n_spp_within_usa, big.mark=",")`) fall within the US waters. These numbers indicate that the lack of geographic coordinates associated with the specimens in the database contribute at least significantly to the discrepency, but that a large fraction of these species have no digital records of their occurrence in the Gulf of Mexico. In other words, additional digitization of collections (including capturing their geographic coordinates information) are needed. For instance, one can expect that the digitization coverage for the Gulf of Mexico will greatly increase once records from the Florida Wildlife Research Institute collections (St Petersburg, FL) become integrated into the biodiversity information systems.

@@

We investigated the proportion of species discovered since the publication of the faunistic lists using iDigBio records. They represent `r sum(recently_collected$gom_in_idig$min_year > 2001)` species for the Gulf of Mexico, and `r sum(recently_collected$pnw_in_idig$min_year > 1987)` species for the Pacific Northwest. While errors in geographic coordinates cannot be completely ruled out, it is more likely that the main contributing factor was that the authors of these checklists were not aware of the presence of these species based only on museum records, as these lists were mostly assembled from publications, and only a small fraction of museum records are used in publications. However, museum specimens are critical to validate the occurrence of species [@Rocha2014], especially for marine invertebrates for which species limits based on morphology have been known to be challenged by other lines of evidence including molecular data [@Knowlton2000].

@@

We combined `r format(nrow(idigbio_records), big.mark=",")` museum records for marine invertebrates collected within the EEZ of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r format(length(unique(idigbio_records$worms_valid_name)), big.mark=",")` species. Most samples were collected on the continental shelf (Fig. \@ref(fig:map-diversity-sampling)A). For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline. For the West coast, the Pacific Northwest, the San Francisco Bay Area, and Southern California have the highest richness (Fig. \@ref(fig:map-diversity-sampling)B).

@@

These patterns of spatial distribution of the species richness are strongly confounded by the sampling effort. When investigating the relation between the number of samples and the number of species for each cell of the map (Fig. \@ref(fig:correlation-sampling)), it takes about 100 records per cell before there is some redundancy in the sampled species (each record represent a different species).

@@

The inadequacy of the sampling effort compared to the diversity of the fauna is also reflected by the fact that `r prop_singleton_species`% of the species are known from a single record. The strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new observations and new species (Fig. \@ref(fig:correlation-sampling)). Another indication of this inadequacy is the relation between the number of collected specimen and the number of species recorded through time (Fig. \@ref(fig:idigbio-through-time)). Initial explorations in the late ninetieth century lead to an initial exponential growth in the knowledge of the fauna, that tapered off, until the 1970's when intensive collecting efforts led to an increase in the recording of new taxa (Fig. \@ref(fig:idigbio-through-time)). Since then, the slowing down of collections after the 1970s has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000 (`r format_output(n_spp_comparison$n_spp_2017_2000/18)` species.year^-1^), while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970 (`r format_output(n_spp_comparison$n_spp_1970_1960/11)` species.year^-1^). Therefore, coordinated large-scale biodiversity expeditions will continue to generate a wealth of data and new records. For instance, a recent study in Western Australia reported that 135 samples led to 805 species of crustaceans and polychaetes, 94.6% and 72% of which were undescribed [@Poore2014].

```{r idigbio-through-time, fig.cap='Cumulative number of samples (A) and species (B) of marine invertebrates recorded through time based on iDigBio records. Relation between number of samples collected, and number of species recorded for the first time (C). (\\#fig:idigbio-through-time)'}
knitr::include_graphics("figures/idigbio_through_time.pdf")
```

@@

To assess the status of taxonomic expertise in the US, we calculated the proportion of specimens identified at the species level through time. For all taxa, excepts for molluscs, there is a consistent decline, indicating that specimens remain for the most part unidentified in collections (Fig.  \@ref(fig:identification-through-time)). For diverse groups such as annelids or arthropods, the proportion of specimens identified at the species-level is now below 30%. The number of scientists describing new species might be increasing globally [@Costello2013], but local experts are needed to keep identifying the species that get collected.
